---
title: "RNAseq"
author: "Shira, Moria, Naama"
date: "2023-06-22"
output: html_document
---
## Libraries

```{r setup, include=FALSE}
library(googledrive)
library(corrplot)
library(ggplot2)
library(randomForest)
library(caret)
library(readr)
library(class)
library(RColorBrewer)
library(FNN)
library(cluster)
library(survival)
library(survminer)
library(ggpubr)
library(dplyr)
library(httr)
```

## For first time, please uncomment the following code and run:

```{r}

# If its the first time runing the code on your devise: 
drive_auth()

# clinical data
url="https://drive.google.com/file/d/1sIeLwzRQvszjkUNsefTtVdhS74Y09zbz/view?usp=sharing"
drive_download(url,overwrite=TRUE)

# Rna seq
url="https://drive.google.com/file/d/1Ecqg0itgMpaVxuy3lbQt_OHQm77JQpif/view?usp=sharing"
drive_download(url,overwrite=TRUE)

# mutations
url="https://drive.google.com/file/d/1165vt_jwnRiXLBHmGyC1XAobGGSPQXRt/view?usp=sharing"
drive_download(url,overwrite=TRUE)

url="https://drive.google.com/file/d/1_98OI7FKtb_Ky3lHyh0QwYPt43xYZkF2/view?usp=sharing"
drive_download(url,overwrite=TRUE)


url="https://drive.google.com/file/d/10tB0PeVkCKiTa85e1Ynnq1rTHvt68lGJ/view?usp=sharing"
drive_auth()
Breast_cancer <- read_delim("brca_tcga_clinical_data.tsv", sep = "\t")
```

```{r}
Breast_cancer <- read.delim("brca_tcga_clinical_data.tsv")
```

brca_tcga_clinical_data.tsv
```{r}
bc_data<- read.csv("Breast_cancer_staging_V2.csv")
head(bc_data)
```
scaled_x_rna_seq2.csv
```{r}
scaled_data <- read.csv("scaled_x_rna_seq2.csv")
dim(scaled_data)
head(scaled_data)
```
y_mutations
```{r}
y_mutations <- read.csv("y_mutations2.csv")
dim(y_mutations)
head(y_mutations)
```
bc_data_df
```{r}
head(bc_data$breast_carcinoma_progesterone_receptor_status.x)
  
subtype_df <- subset(bc_data, select =
                       c(bcr_patient_barcode,subtype_BRCA_Subtype_PAM50,breast_carcinoma_estrogen_receptor_status.x,
                         breast_carcinoma_progesterone_receptor_status.x,lab_proc_her2_neu_immunohistochemistry_receptor_status.x))

colnames(subtype_df) <- c("id", "subtype", "estrogen_receptor","progesterone_receptor","her2_receptor")
head(subtype_df)
```

## pie charts

```{r}
er <- table(subtype_df$estrogen_receptor)
myPalette <- brewer.pal(length(er), "Set3") 
pie(er , labels = names(er), border="white", col=myPalette ,main="ER Status")
her2 <- table(subtype_df$her2_receptor)
myPalette <- brewer.pal(length(her2), "Set3") 
pie(her2 , labels = names(her2), border="white", col=myPalette ,main="HER2 Status")
pr <- table(subtype_df$progesterone_receptor)
myPalette <- brewer.pal(length(pr), "Set3") 
pie(pr , labels = names(pr), border="white", col=myPalette ,main="PR Status")
types <- table(subtype_df$subtype)
myPalette <- brewer.pal(length(types), "Set3") 
pie(types , labels = names(types), border="white", col=myPalette ,main="subtype")
```



```{r, warning=FALSE}

subtype_df <- subset(bc_data, select =
                       c(bcr_patient_barcode,subtype_BRCA_Subtype_PAM50,breast_carcinoma_estrogen_receptor_status.x,
                         breast_carcinoma_progesterone_receptor_status.x,lab_proc_her2_neu_immunohistochemistry_receptor_status.x))

colnames(subtype_df) <- c("id", "subtype", "estrogen_receptor","progesterone_receptor","her2_receptor")
head(subtype_df)


race_counts <- table(Breast_cancer$Race.Category)
myPalette <- brewer.pal(length(race_counts), "Set2") 
pie(race_counts , labels = c("NATIVE","ASIAN","AFRICAN AMERICAN", "CAUCASIAN"), border="white", col=myPalette )

sex_counts <- table(Breast_cancer$Sex)
myPalette <- brewer.pal(length(sex_counts), "Set3") 
pie(sex_counts , labels = names(sex_counts), border="white", col=myPalette )

survival_counts <- table(Breast_cancer$Overall.Survival.Status)
myPalette <- brewer.pal(length(survival_counts), "Set2") 
pie(survival_counts , labels = c("ALIVE","DEAD"), border="white", col=myPalette )

Menopause <- table(Breast_cancer$Menopause.Status)
myPalette <- brewer.pal(length(Menopause), "Set3") 
pie(Menopause , labels = c("Indeterminate","Peri: 6-12 mos","Post : >12 mos","Pre"), border="white", col=myPalette )

```

```{r, warning=FALSE}

Breast_cancer %>% 
    group_by(Diagnosis.Age) %>%
    count() %>%
    ggplot() +
    geom_col(aes(Diagnosis.Age,n),fill = 'pink')+
    ggtitle("Age Analysis")+
    xlab("Age")+
    ylab("Age Count")
Age<-(Breast_cancer$Diagnosis.Age)
Mutation_Count<-(Breast_cancer$Mutation.Count)

#plot(Age,Mutation_Count, ylab = "Mutation Count", xlab = "Age", main = "Mutation_Count vs. Age")
```

```{r}
numeric_bc <- Breast_cancer[, c("Mutation.Count", "Fraction.Genome.Altered", "Diagnosis.Age", "Days.to.Sample.Collection.", "Lymph.Node.s..Examined.Number", "Sample.Initial.Weight")]
# Calculate the correlation matrix
cor_numeric_bc <- cor(numeric_bc, use = "pairwise.complete.obs")
# Subset the correlation matrix for variables with missing values
cor_missing_values <- cor_numeric_bc[, colSums(is.na(numeric_bc)) > 0]
corrplot(cor_missing_values, method = "square", type = "upper")
```

Breast_cancer$Neoplasm.American....
Breast_cancer$Patient.Primary.Tumor.Site
Breast_cancer$Patient.Metastatic.Sites
Breast_cancer$Cancer.Type.Detailed
Breast_cancer$Prior.Cancer.Diagnosis.Occurence
Breast_cancer$Surgical.procedure.first
Breast_cancer$Disease.Surgical.Margin.Status
Breast_cancer$Menopause.Status

Breast_cancer$Sex
Breast_cancer$Race.Category

## survival plots
```{r}
data_bc_filtered <- Breast_cancer
data_bc_filtered$Vital_status <- ifelse(data_bc_filtered$Overall.Survival.Status == "0:LIVING", 0, 1)

# Specify the correct column names for time and status in the Surv() function
fit <- survfit(Surv(Overall.Survival..Months., Vital_status) ~ Sex, data = data_bc_filtered)

# Create a new plot with survival curves for high and low expression
g <- ggsurvplot(fit, data = data_bc_filtered, risk.table = FALSE, pval = TRUE,
                risk.table.y.text = FALSE, risk.table.height = 0.3,
                palette = c("#2E9FDF", "purple"), conf.int = TRUE,
                xlab = "Time from Diagnosis (months)",
                ylab = "Proportion Survival",
                font.x = c(17, "bold", "black"),
                font.y = c(17, "bold", "black"),
                font.tickslab = c(17, "bold", "black"),
                legend = c(0.89, 0.89),
                title = "Survival\n")

# Display the survival plot
g

```

```{r}
tmp <- y_mutations[, c("X", "TP53")]
tmp2<-Breast_cancer[,c("Patient.ID","Overall.Survival.Status","Overall.Survival..Months.")]
tmp2$Overall.Survival.Status<- ifelse(tmp2$Overall.Survival.Status == "0:LIVING", 0, 1)
tail(tmp2)
colnames(tmp) <- c("Patient.ID", "TP53")
head(tmp)
```

```{r}
tp53_data<-(merge(tmp2, tmp, by= "Patient.ID"))

# Specify the correct column names for time and status in the Surv() function
fit <- survfit(Surv(Overall.Survival..Months., Overall.Survival.Status) ~ TP53, data = tp53_data)

# Create a new plot with survival curves for high and low expression
g <- ggsurvplot(fit, data = tp53_data, risk.table = FALSE, pval = TRUE,
                risk.table.y.text = FALSE, risk.table.height = 0.3,
                palette = c("#2E9FDF", "purple"), conf.int = TRUE,
                xlab = "Time from Diagnosis (months)",
                ylab = "Proportion Survival",
                font.x = c(17, "bold", "black"),
                font.y = c(17, "bold", "black"),
                font.tickslab = c(17, "bold", "black"),
                legend = c(0.89, 0.89),
                title = "Survival\n")

# Display the survival plot
g

```

```{r}
tmp <- subtype_df[, c("id", "subtype")]
tmp2<-Breast_cancer[,c("Patient.ID","Overall.Survival.Status","Overall.Survival..Months.")]
tmp2$Overall.Survival.Status<- ifelse(tmp2$Overall.Survival.Status == "0:LIVING", 0, 1)

head(tmp)
colnames(tmp2) <- c("id", "Survival_Status", "Survival_Months")
subtype_alive_data<-(merge(tmp2, tmp, by= "id"))
head(subtype_alive_data)
# Specify the correct column names for time and status in the Surv() function
fit <- survfit(Surv(Survival_Months, Survival_Status) ~ subtype, data = subtype_alive_data)

# Create a new plot with survival curves for high and low expression
g <- ggsurvplot(fit, data = subtype_alive_data, risk.table = FALSE, pval = TRUE,
                risk.table.y.text = FALSE, risk.table.height = 0.3,
                palette = c("#2E9FDF", "purple", "green", "yellow","red"), conf.int = TRUE,
                xlab = "Time from Diagnosis (months)",
                ylab = "Proportion Survival",
                font.x = c(17, "bold", "black"),
                font.y = c(17, "bold", "black"),
                font.tickslab = c(17, "bold", "black"),
                legend = c(0.89, 0.89),
                title = "Survival\n")

# Display the survival plot
g
```

```{r}
RNAseq_data_survival <- merge(scaled_data, subset(Breast_cancer, select = c(Patient.ID,Overall.Survival.Status)), by.x = "X", by.y = "Patient.ID")
RNAseq_data_survival <- na.omit(RNAseq_data_survival)

head(RNAseq_data_survival)
data_rna_survival <- RNAseq_data_survival[,2:56500]
head(colnames(data_rna_survival))
```

## PCA on y_mutation data and scaled rna seq
mereged mutation data
```{r}
bc_data_df <- subset(subtype_df, select = colnames(subtype_df))
head(bc_data_df)
mutation_data <- merge(y_mutations, bc_data_df, by.x = "X", by.y = "id")
head(mutation_data)
data_x_y<-merge(mutation_data, scaled_data, by.x = "X", by.y = "X")
```

```{r}
tail(colnames(data_x_y))
data_x_y_pca <- data_x_y[, !(names(data_x_y) %in% c("X","subtype","estrogen_receptor","progesterone_receptor","her2_receptor"))]
```

```{r}
standardized_data_XY <- scale(data_x_y_pca[, -ncol(data_x_y_pca)])
pca_result_XY <- prcomp(standardized_data_XY)
pc_scores_XY <- as.data.frame(pca_result_XY$x)
```

```{r}
tail(colnames(mutation_data))
pc_scores_XY$subtype <- factor(mutation_data$subtype)
pc_scores_XY$ER.Status.By.IHC <- factor(mutation_data$estrogen_receptor)
pc_scores_XY$IHC.HER2 <- factor(mutation_data$her2_receptor)
pc_scores_XY$PR.status.by.ihc <- factor(mutation_data$progesterone_receptor)
```


```{r}
ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = subtype)) +
  geom_point() +
  labs(title = "PCA Plot - subtype") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = ER.Status.By.IHC)) +
  geom_point() +
  labs(title = "PCA Plot - ER Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = IHC.HER2)) +
  geom_point() +
  labs(title = "PCA Plot - IHC HER2") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = PR.status.by.ihc)) +
  geom_point() +
  labs(title = "PCA Plot - PR Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```

## PCA - on mutation
```{r}
bc_data_df <- subset(subtype_df, select = colnames(subtype_df))
head(bc_data_df)
mutation_data <- merge(y_mutations, bc_data_df, by.x = "X", by.y = "id")
head(mutation_data)
```

```{r}
pca_data_y <- mutation_data[, c(colnames(mutation_data)[2:45])]
standardized_data_y <- scale(pca_data_y[, -ncol(pca_data_y)])
pca_result_y <- prcomp(standardized_data_y)
pc_scores_y <- as.data.frame(pca_result_y$x)
```

```{r}
pc_scores_y$subtype <- factor(mutation_data$subtype)
pc_scores_y$estrogen_receptor <- factor(mutation_data$estrogen_receptor)
pc_scores_y$her2_receptor <- factor(mutation_data$her2_receptor)
pc_scores_y$progesterone_receptor <- factor(mutation_data$progesterone_receptor)
```

```{r}
ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = subtype)) +
  geom_point() +
  labs(title = "PCA Plot - Overall Survival Status") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = estrogen_receptor)) +
  geom_point() +
  labs(title = "PCA Plot - ER Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = her2_receptor)) +
  geom_point() +
  labs(title = "PCA Plot - IHC HER2") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = progesterone_receptor)) +
  geom_point() +
  labs(title = "PCA Plot - PR Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```

```{r}
gene_summaries <- colSums(y_mutations[, -1])
gene_summaries

id_summaries <- rowSums(y_mutations[, -1])
# Create the bar plot
barplot(table(id_summaries), main = "Frequency", xlab = "Number of mutation", ylab = "Count")
```
```{r}
sub_rna <- subset(subtype_df, select = c(id,subtype))
head(sub_rna)

RNAseq_data_subtype <- merge(scaled_data, sub_rna, by.x = "X", by.y = "id")

RNAseq_data_subtype <- na.omit(RNAseq_data_subtype)

rna_data_rf<- RNAseq_data_subtype[,2:56499]
subtype_status<-factor(RNAseq_data_subtype$subtype)

```
```{r}

unique((RNAseq_data_subtype$subtype))
rna_matrix <- as.matrix(rna_data_rf)

# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 4 # Number of clusters
kmeans_result <- kmeans(rna_matrix, centers = k)

# Get the cluster assignments for each data point
cluster_labels <- kmeans_result$cluster

# Assuming 'Class' column is a factor
class_labels <- RNAseq_data_subtype$subtype

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results, cluster labels, and class labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels), Class = factor(class_labels))

# Plot PCA with clusters and class labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = Class, shape = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using k-means (k = 4)") +
  #scale_color_manual(values = class_colors) +
  guides(shape = guide_legend(title = "Cluster"), color = guide_legend(title = "Class"))
```
## knn and random forest on subtype
```{r}
# Perform k-NN clustering with 4 clusters

unique((RNAseq_data_subtype$subtype))
rna_matrix <- as.matrix(rna_data_rf)
class_labels <- RNAseq_data_subtype$subtype
# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 4 # Number of clusters
#kmeans_result <- kmeans(rna_matrix, centers = k)
knn_result <- knn(rna_matrix, rna_matrix, cl = class_labels, k = k)
# Get the cluster assignments for each data point

cluster_labels <- knn_result
# Assuming 'Class' column is a factor
class_labels <- RNAseq_data_subtype$subtype

library(ggplot2)

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results, cluster labels, and class labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels), Class = factor(class_labels))

# Plot PCA with clusters and class labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = Class, shape = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using knn (k = 4)") +
  #scale_color_manual(values = class_colors) +
  guides(shape = guide_legend(title = "Cluster"), color = guide_legend(title = "Class"))
```

```{r}
# Convert class labels to factor if they are not already
class_labels <- factor(RNAseq_data_subtype$subtype)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_indices <- sample(1:nrow(rna_matrix), nrow(rna_matrix) * 0.7)  # 70% for training
train_data <- rna_matrix[train_indices, ]
train_labels <- class_labels[train_indices]
test_data <- rna_matrix[-train_indices, ]
test_labels <- class_labels[-train_indices]

# Train the Random Forest model
rf_model <- randomForest(train_data, train_labels, ntree = 100)

# Make predictions on the test data
rf_predictions <- predict(rf_model, test_data)

# Calculate the accuracy of the model
accuracy <- sum(rf_predictions == test_labels) / length(test_labels)
print(paste("Accuracy:", accuracy))

# Plot the confusion matrix
table(test_labels, rf_predictions)
```
## ER kmeans with rna seq 
```{r}
#subtype_df$
sub_rna <- subset(subtype_df, select = c(id,estrogen_receptor))
head(sub_rna)

RNAseq_data_er <- merge(scaled_data, sub_rna, by.x = "X", by.y = "id")
#RNAseq_data_er <- as.data.frame(RNAseq_data_er)

RNAseq_data_er <- na.omit(RNAseq_data_er)


head(RNAseq_data_er)
rna_data_rf<- RNAseq_data_er[,2:56499]
er_status <- RNAseq_data_er$estrogen_receptor
er_status <- factor(er_status)

```

```{r}
rna_matrix <- as.matrix(rna_data_rf)

# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 3  # Number of clusters
kmeans_result <- kmeans(rna_matrix, centers = k)

# Get the cluster assignments for each data point
cluster_labels <- kmeans_result$cluster

# Assuming 'Class' column is a factor
class_labels <- RNAseq_data_er$estrogen_receptor

# Visualize the clusters and class labels (optional)
# Example: PCA plot

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results, cluster labels, and class labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels), Class = factor(class_labels))

# Define colors for the class labels
class_colors <- c("Positive" = "blue", "Negative" = "red", "NA" = "green", "Indeterminate" = "orange")

# Plot PCA with clusters and class labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = Class, shape = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using k-means (k = 3)") +
  scale_color_manual(values = class_colors) +
  guides(shape = guide_legend(title = "Cluster"), color = guide_legend(title = "Class"))

```

## PR kmeans and rna seq 
```{r}
sub_rna <- subset(subtype_df, select = c(id,progesterone_receptor))
head(sub_rna)

RNAseq_data_pr <- merge(scaled_data, sub_rna, by.x = "X", by.y = "id")

RNAseq_data_pr <- na.omit(RNAseq_data_pr)

head(RNAseq_data_pr)
rna_data_rf<- RNAseq_data_pr[,2:56499]
pr_status <- RNAseq_data_pr$progesterone_receptor
pr_status <- factor(pr_status)
```

```{r}
rna_matrix <- as.matrix(rna_data_rf)

# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 3  # Number of clusters
kmeans_result <- kmeans(rna_matrix, centers = k)

# Get the cluster assignments for each data point
cluster_labels <- kmeans_result$cluster

# Assuming 'Class' column is a factor
class_labels <- RNAseq_data_pr$progesterone_receptor

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results, cluster labels, and class labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels), Class = factor(class_labels))

# Define colors for the class labels
class_colors <- c("Positive" = "blue", "Negative" = "red", "NA" = "green", "Indeterminate" = "orange")

# Plot PCA with clusters and class labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = Class, shape = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using k-means (k = 3)") +
  scale_color_manual(values = class_colors) +
  guides(shape = guide_legend(title = "Cluster"), color = guide_legend(title = "Class"))

```
