---
title: "project ml in r Breast cancer"
author: "Shira, Moria, Naama"
date: "2023-08-13"
output: html_document
---

```{r setup, include=FALSE}
library(googledrive)
library(glmnetUtils)
library(corrplot)
library(randomForest)
library("devtools")
library("factoextra")
library(ggplot2)
library(rpart)
library(e1071)
library(randomForest)
library(caret)
library(readr)
library(class)
library(RColorBrewer)
library(FNN)
library(cluster)
library(survival)
library(survminer)
library(ggpubr)
library(dplyr)
library(httr)
```


```{r}
y_mutations <- read.csv("y_mutations2.csv")
head(y_mutations)
```

```{r}
Breast_cancer <- read.delim("brca_tcga_clinical_data.tsv")
head(Breast_cancer)
```

brca_tcga_clinical_data.tsv
```{r}
bc_data<- read.csv("Breast_cancer_staging_V2.csv")
head(bc_data)
```
another data for comparezition
```{r}
df = read.table("data_mrna_illumina_microarray_zscores_ref_diploid_samples.txt",header=TRUE)
df <- df %>%   select(-Entrez_Gene_Id)

df <- t(df)
# Convert the transposed matrix to a data frame
df <- as.data.frame(df, stringsAsFactors = FALSE)

# Extract header
header <- df[1, ]

# Assign headers to the data frame
colnames(df) <- header
df <- df[-1, ]
colnames(df)[0] <- "PATIENT_ID"

head(df)
dim(df)

clinical_df=read.table("data_clinical_patient.txt",sep='\t',header=TRUE)
clinical_df$PATIENT_ID <- gsub("-", ".", clinical_df$PATIENT_ID)
head(clinical_df)
dim(clinical_df)
```

```{r}
clinical_df_subset <- subset(clinical_df, select =
                       c(PATIENT_ID,HER2_SNP6,CLAUDIN_SUBTYPE,ER_IHC,INFERRED_MENOPAUSAL_STATE))

clinical_df_subset <- na.omit(clinical_df_subset)
row.names(clinical_df_subset) <- clinical_df_subset$PATIENT_ID
clinical_df_subset <- clinical_df_subset[, -1]

head(clinical_df_subset)
```

```{r}
merged_df <- merge(clinical_df_subset, df, by = 0, all = TRUE)
rownames(merged_df) <- merged_df$Row.names
merged_df <- merged_df[-1]
merged_df<-na.omit(merged_df)
```

```{r}
head(merged_df)
dim(merged_df)
```

```{r}
subtype_data_df <-subset(merged_df, select = -c(HER2_SNP6,ER_IHC,INFERRED_MENOPAUSAL_STATE))
her2_data_df <- subset(merged_df, select = -c(CLAUDIN_SUBTYPE,ER_IHC,INFERRED_MENOPAUSAL_STATE))
estrogen_data_df <- subset(merged_df, select = -c(CLAUDIN_SUBTYPE,INFERRED_MENOPAUSAL_STATE,HER2_SNP6))
menapausal_data_df <-subset(merged_df, select = -c(CLAUDIN_SUBTYPE,ER_IHC,HER2_SNP6))
numeric_data_df <-as.data.frame(lapply(subset(merged_df, select = -c(CLAUDIN_SUBTYPE,ER_IHC,HER2_SNP6,INFERRED_MENOPAUSAL_STATE)), as.numeric))

```


scaled_x_rna_seq2.csv
```{r}
scaled_data <- read.csv("scaled_x_rna_seq2.csv")
dim(scaled_data)
head(scaled_data)
```

```{r}
subtype_df <- subset(bc_data, select =
                       c(bcr_patient_barcode,subtype_BRCA_Subtype_PAM50,breast_carcinoma_estrogen_receptor_status.x,
                         breast_carcinoma_progesterone_receptor_status.x,lab_proc_her2_neu_immunohistochemistry_receptor_status.x))

colnames(subtype_df) <- c("id", "subtype", "estrogen_receptor","progesterone_receptor","her2_receptor")
head(subtype_df)

# Remove rows with NA values from the RNA-seq data
subtype_df <- na.omit(subtype_df)
dim(subtype_df)
```

```{r}
subtype_data <- subset(subtype_df, select = c(id,subtype))
head(subtype_data)

her2_data <- subset(subtype_df, select = c(id,her2_receptor))
head(her2_data)

progesterone_data <- subset(subtype_df, select = c(id,progesterone_receptor))
head(progesterone_data)

estrogen_data <- subset(subtype_df, select = c(id,estrogen_receptor))
head(estrogen_data)


y_mutations<-na.omit(y_mutations)
scaled_data<-na.omit(scaled_data)
     
y_mutations_data_subtype <- merge(y_mutations, subtype_data, by.x = "X", by.y = "id")
y_mutations_data_her2 <- merge(y_mutations, her2_data, by.x = "X", by.y = "id")
y_mutations_data_progesterone <- merge(y_mutations, progesterone_data, by.x = "X", by.y = "id")
y_mutations_data_estrogen <- merge(y_mutations, estrogen_data, by.x = "X", by.y = "id")

RNAseq_data_subtype <- merge(scaled_data, subtype_data, by.x = "X", by.y = "id")
RNAseq_data_her2 <- merge(scaled_data, her2_data, by.x = "X", by.y = "id")
RNAseq_data_progesterone <- merge(scaled_data, progesterone_data, by.x = "X", by.y = "id")
RNAseq_data_estrogen <- merge(scaled_data, estrogen_data, by.x = "X", by.y = "id")
```
```{r}
data<-merge(scaled_data, subtype_df, by.x = "X", by.y = "id")
numeric_data<-(subset(data, select = -c(X, subtype,estrogen_receptor,progesterone_receptor,her2_receptor)))
head(numeric_data)
```

##pca 
This analysis can be insightful in understanding the variation in the RNA-seq data and how it relates to the different receptor subtypes.

1) pca on the second data set
```{r}
head(numeric_data_df)
pca <- prcomp(numeric_data_df)
```

```{r}
fviz_pca_ind(pca, col.ind="cos2", geom = "point") +
      scale_color_gradient2(low="white", mid="red",high="blue", midpoint=0.4)+ theme_minimal()
```

```{r}
# Control the transparency of the color by the contributions
fviz_pca_ind(pca, col.ind="contrib") +
      scale_color_gradient2(low="red", mid="blue",
      high="red", midpoint=0.4)
```

```{r,warning=FALSE}
p <- fviz_pca_ind(pca, label="none", habillage=her2_data_df$HER2_SNP6,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r}
p <- fviz_pca_ind(pca, label="none", habillage=subtype_data_df$CLAUDIN_SUBTYPE,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r}
p <- fviz_pca_ind(pca, label="none", habillage=estrogen_data_df$ER_IHC,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r}
p <- fviz_pca_ind(pca, label="none", habillage=menapausal_data_df$INFERRED_MENOPAUSAL_STATE,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

##pca 
This analysis can be insightful in understanding the variation in the RNA-seq data and how it relates to the different receptor subtypes.
2) on the first data set
```{r}
pca <- prcomp(numeric_data)
```

```{r}
fviz_pca_ind(pca, col.ind="cos2", geom = "point") +
      scale_color_gradient2(low="white", mid="red",high="blue", midpoint=0.2)+ theme_minimal()
```

```{r}
# Control the transparency of the color by the contributions
fviz_pca_ind(pca, col.ind="contrib") +
      scale_color_gradient2(low="lightblue", mid="blue",
      high="red", midpoint=1)
```

```{r,warning=FALSE}
p <- fviz_pca_ind(pca, label="none", habillage=RNAseq_data_subtype$subtype,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r}
p <- fviz_pca_ind(pca, label="none", habillage=RNAseq_data_her2$her2_receptor,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r,warning=FALSE}
p <- fviz_pca_ind(pca, label="none", habillage=RNAseq_data_estrogen$estrogen_receptor,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```

```{r,warning=FALSE,message=FALSE}
p <- fviz_pca_ind(pca, label="none", habillage=RNAseq_data_progesterone$progesterone_receptor,
             addEllipses=TRUE, ellipse.level=0.95)
print(p)
```


```{r}
pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(RNAseq_data_estrogen$estrogen_receptor)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()

pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(RNAseq_data_her2$her2_receptor)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()

pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(RNAseq_data_progesterone$progesterone_receptor)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()


pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(RNAseq_data_subtype$subtype)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()

```

```{r}
data<-merge(scaled_data, subtype_df, by.x = "X", by.y = "id")
numeric_data<-(subset(data, select = -c(X, subtype,estrogen_receptor,progesterone_receptor,her2_receptor)))
head(numeric_data)
```

data without undefined her2 status
```{r}
filtered_rna_data <- data %>%
  filter(her2_receptor %in% c("Positive", "Negative"))
numeric_data_filter<-(subset(filtered_rna_data, select = -c(X, subtype,estrogen_receptor,progesterone_receptor,her2_receptor)))

variance_threshold <- 0.15# Set the variance threshold
gene_variances <- apply(numeric_data, 2, var)# Calculate the variance for each gene
selected_genes <- names(gene_variances)[gene_variances >= variance_threshold]# Filter genes based on the variance threshold
selected_rna_data <- numeric_data %>% select(all_of(selected_genes))
dim(numeric_data_filter)
dim(selected_rna_data)
variance_threshold <- 0.15
gene_variances <- apply(numeric_data_filter, 2, var)
selected_genes <- names(gene_variances)[gene_variances >= variance_threshold]
selected_rna_data <- numeric_data_filter %>%   select(all_of(selected_genes))
```

```{r}
set.seed(123)
data_indices <- seq_len(nrow(selected_rna_data))
train_indices <- sample(data_indices, size = 0.7 * nrow(selected_rna_data))
train_data <- selected_rna_data[train_indices, ]
test_data <- selected_rna_data[-train_indices, ]
train_labels_her2 <- as.factor(filtered_rna_data$her2_receptor[train_indices])
test_labels_her2 <- filtered_rna_data$her2_receptor[-train_indices]
```
her 2 svm model (positive or negative)
```{r}
svm_model_her2 <- svm(x = train_data, y = train_labels_her2, kernel = "radial")

predictions_her2 <- predict(svm_model_her2, newdata = test_data)

accuracy_her2 <- mean(predictions_her2 == test_labels_her2)

print(paste("Accuracy for HER2 Receptor:", accuracy_her2))
```

# ml models
```{r}
set.seed(123)

dim(numeric_data)

data_indices <- seq_len(nrow(numeric_data))

# Sample 70% of the row indices for the training set
train_indices <- sample(data_indices, size = 0.7 * nrow(numeric_data))

# Create the training set using the sampled row indices
train_data <- numeric_data[train_indices, ]

# Create the testing set by excluding the training set row indices
test_data <- numeric_data[-train_indices, ]

train_labels_er <- as.factor(data$estrogen_receptor[train_indices])
test_labels_er <- data$estrogen_receptor[-train_indices]

train_labels_pr <- as.factor(data$progesterone_receptor[train_indices])
test_labels_pr <- data$progesterone_receptor[-train_indices]

train_labels_her2 <- as.factor(data$her2_receptor[train_indices])
test_labels_her2 <- data$her2_receptor[-train_indices]

train_labels_subtype <- as.factor(data$subtype[train_indices])
test_labels_subtype <- data$subtype[-train_indices]
```

##svm
```{r}
# Install and load the required package
library(e1071)

# SVM function with radial basis kernel (you can try other kernels as well)
svm_model_er <- svm(x = train_data, y = train_labels_er, kernel = "radial")
svm_model_pr <- svm(x = train_data, y = train_labels_pr, kernel = "radial")
svm_model_her2 <- svm(x = train_data, y = train_labels_her2, kernel = "radial")
svm_model_subtype <- svm(x = train_data, y = train_labels_subtype, kernel = "radial")

# Make predictions on the testing set
predictions_er <- predict(svm_model_er, newdata = test_data)
predictions_pr <- predict(svm_model_pr, newdata = test_data)
predictions_her2 <- predict(svm_model_her2, newdata = test_data)
predictions_subtype <- predict(svm_model_subtype, newdata = test_data)

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_pr <- mean(predictions_pr == test_labels_pr)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for Progesterone Receptor:", accuracy_pr))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2)) #"Accuracy for HER2 Receptor: 0.798305084745763"
print(paste("Accuracy for Subtype:", accuracy_subtype))

```
##decision tree
```{r}
variance_threshold <- 0.15
gene_variances <- apply(numeric_data, 2, var)# Calculate the variance for each gene
selected_genes <- names(gene_variances)[gene_variances >= variance_threshold]# Filter genes based on the variance threshold
selected_rna_data <- numeric_data %>% select(all_of(selected_genes))# Create a new tibble with selected genes
```

```{r}
set.seed(123)

head(selected_rna_data)
dim(selected_rna_data)

data_indices <- seq_len(nrow(selected_rna_data))

# Sample 70% of the row indices for the training set
train_indices <- sample(data_indices, size = 0.7 * nrow(selected_rna_data))

# Create the training set using the sampled row indices
train_data <- selected_rna_data[train_indices, ]

# Create the testing set by excluding the training set row indices
test_data <- selected_rna_data[-train_indices, ]

train_labels_her2 <- as.factor(data$her2_receptor[train_indices])
test_labels_her2 <- data$her2_receptor[-train_indices]

train_labels_er <- as.factor(data$estrogen_receptor[train_indices])
test_labels_er <- data$estrogen_receptor[-train_indices]

train_labels_pr <- as.factor(data$progesterone_receptor[train_indices])
test_labels_pr <- data$progesterone_receptor[-train_indices]

train_labels_subtype <- as.factor(data$subtype[train_indices])
test_labels_subtype <- data$subtype[-train_indices]

```

```{r}
# Decision tree for Estrogen Receptor
tree_er <- rpart(factor(train_labels_er) ~ ., data = train_data, method = "class")

# Decision tree for Progesterone Receptor
tree_pr <- rpart(factor(train_labels_pr) ~ ., data = train_data, method = "class")

# Decision tree for HER2 Receptor
tree_her2 <- rpart(factor(train_labels_her2) ~ ., data = train_data, method = "class")

# Decision tree for Subtype
tree_subtype <- rpart(factor(train_labels_subtype) ~ ., data = train_data, method = "class")

# Make predictions on the testing set
predictions_er <- predict(tree_er, newdata = test_data, type = "class")
predictions_pr <- predict(tree_pr, newdata = test_data, type = "class")
predictions_her2 <- predict(tree_her2, newdata = test_data, type = "class")
predictions_subtype <- predict(tree_subtype, newdata = test_data, type = "class")

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_pr <- mean(predictions_pr == test_labels_pr)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for Progesterone Receptor:", accuracy_pr))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2))
print(paste("Accuracy for Subtype:", accuracy_subtype))

```
#randomForest
```{r}
num_trees <- 100

# Random Forest for Estrogen Receptor
rf_er <- randomForest(factor(train_labels_er) ~ ., data = train_data, ntree = num_trees)

# Random Forest for Progesterone Receptor
rf_pr <- randomForest(factor(train_labels_pr) ~ ., data = train_data, ntree = num_trees)

# Random Forest for HER2 Receptor
rf_her2 <- randomForest(factor(train_labels_her2) ~ ., data = train_data, ntree = num_trees)

# Random Forest for Subtype
rf_subtype <- randomForest(factor(train_labels_subtype) ~ ., data = train_data, ntree = num_trees)

# Make predictions on the testing set
predictions_er <- predict(rf_er, newdata = test_data)
predictions_pr <- predict(rf_pr, newdata = test_data)
predictions_her2 <- predict(rf_her2, newdata = test_data)
predictions_subtype <- predict(rf_subtype, newdata = test_data)

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_pr <- mean(predictions_pr == test_labels_pr)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for Progesterone Receptor:", accuracy_pr))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2))
print(paste("Accuracy for Subtype:", accuracy_subtype))

```



### ml alg on the second data set

subtype_data_df <-subset(merged_df, select = -c(HER2_SNP6,ER_IHC,INFERRED_MENOPAUSAL_STATE))
her2_data_df <- subset(merged_df, select = -c(CLAUDIN_SUBTYPE,ER_IHC,INFERRED_MENOPAUSAL_STATE))
estrogen_data_df <- subset(merged_df, select = -c(CLAUDIN_SUBTYPE,INFERRED_MENOPAUSAL_STATE,HER2_SNP6))
numeric_data_df <-as.data.frame(lapply(subset(merged_df, select = -c(CLAUDIN_SUBTYPE,ER_IHC,HER2_SNP6,INFERRED_MENOPAUSAL_STATE)), as.numeric))



```{r}
set.seed(123)

dim(numeric_data_df)
#dim(selected_numeric_data_df)

data_indices <- seq_len(nrow(selected_numeric_data_df))

# Sample 70% of the row indices for the training set
train_indices <- sample(data_indices, size = 0.7 * nrow(selected_numeric_data_df))

# Create the training set using the sampled row indices
train_data <- selected_numeric_data_df[train_indices, ]

# Create the testing set by excluding the training set row indices
test_data <- selected_numeric_data_df[-train_indices, ]

train_labels_er <- as.factor(estrogen_data_df$ER_IHC[train_indices])
test_labels_er <- estrogen_data_df$ER_IHC[-train_indices]

train_labels_her2 <- as.factor(her2_data_df$HER2_SNP6[train_indices])
test_labels_her2 <- her2_data_df$HER2_SNP6[-train_indices]

train_labels_subtype <- as.factor(subtype_data_df$CLAUDIN_SUBTYPE[train_indices])
test_labels_subtype <- subtype_data_df$CLAUDIN_SUBTYPE[-train_indices]
```
```{r}
svm_model_er <- svm(x = train_data, y = train_labels_er, kernel = "radial")
svm_model_her2 <- svm(x = train_data, y = train_labels_her2, kernel = "radial")
svm_model_subtype <- svm(x = train_data, y = train_labels_subtype, kernel = "radial")

# Make predictions on the testing set
predictions_er <- predict(svm_model_er, newdata = test_data)
predictions_her2 <- predict(svm_model_her2, newdata = test_data)
predictions_subtype <- predict(svm_model_subtype, newdata = test_data)

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2)) #"Accuracy for HER2 Receptor: 0.798305084745763"
print(paste("Accuracy for Subtype:", accuracy_subtype))


```

```{r}
# Decision tree for Estrogen Receptor
tree_er <- rpart(factor(train_labels_er) ~ ., data = train_data, method = "class")

# Decision tree for HER2 Receptor
tree_her2 <- rpart(factor(train_labels_her2) ~ ., data = train_data, method = "class")

# Decision tree for Subtype
tree_subtype <- rpart(factor(train_labels_subtype) ~ ., data = train_data, method = "class")

# Make predictions on the testing set
predictions_er <- predict(tree_er, newdata = test_data, type = "class")
predictions_pr <- predict(tree_pr, newdata = test_data, type = "class")
predictions_her2 <- predict(tree_her2, newdata = test_data, type = "class")
predictions_subtype <- predict(tree_subtype, newdata = test_data, type = "class")

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2))
print(paste("Accuracy for Subtype:", accuracy_subtype))

```
```{r}
library(gmodels)
library(C50)

# apply model in training data (17th column is the label to be predicted)
model <- C5.0(train_data, train_labels_er)
model
pred <- predict(model, test_data)

summary(model)
CrossTable(test_labels_er, pred, prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE)
```

c50_model <- C5.0(train_data, train_labels)

# Make predictions
c50_predictions <- predict(c50_model, newdata = test_data)

# Calculate accuracy
accuracy <- sum(c50_predictions == test_labels) / length(test_labels)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

#randomForest


```{r}
num_trees <- 100

# Random Forest for Estrogen Receptor
rf_er <- randomForest(factor(train_labels_er) ~ ., data = train_data, ntree = num_trees)

# Random Forest for Progesterone Receptor
rf_pr <- randomForest(factor(train_labels_pr) ~ ., data = train_data, ntree = num_trees)

# Random Forest for HER2 Receptor
rf_her2 <- randomForest(factor(train_labels_her2) ~ ., data = train_data, ntree = num_trees)

# Random Forest for Subtype
rf_subtype <- randomForest(factor(train_labels_subtype) ~ ., data = train_data, ntree = num_trees)

# Make predictions on the testing set
predictions_er <- predict(rf_er, newdata = test_data)
predictions_pr <- predict(rf_pr, newdata = test_data)
predictions_her2 <- predict(rf_her2, newdata = test_data)
predictions_subtype <- predict(rf_subtype, newdata = test_data)

# Evaluate the model for each classification task (you can use accuracy, confusion matrix, etc.)
accuracy_er <- mean(predictions_er == test_labels_er)
accuracy_pr <- mean(predictions_pr == test_labels_pr)
accuracy_her2 <- mean(predictions_her2 == test_labels_her2)
accuracy_subtype <- mean(predictions_subtype == test_labels_subtype)

# Print the accuracies
print(paste("Accuracy for Estrogen Receptor:", accuracy_er))
print(paste("Accuracy for Progesterone Receptor:", accuracy_pr))
print(paste("Accuracy for HER2 Receptor:", accuracy_her2))
print(paste("Accuracy for Subtype:", accuracy_subtype))

```


##feature selection

```{r}
library(xgboost)

# Convert data to DMatrix format
if (is.list(train_data)) {
  train_data <- as.matrix(train_data)  # or as.data.frame(train_data)
}

# Convert train_labels_subtype to a numeric vector if needed
if (!is.numeric(train_labels_subtype)) {
  train_labels_subtype <- as.numeric(train_labels_subtype)
}

if (is.list(test_data)) {
  test_data <- as.matrix(test_data)
}

dtrain <- xgb.DMatrix(data = train_data, label = train_labels_subtype)
dtest <- xgb.DMatrix(data = test_data)

# Train an XGBoost model
xgb_model <- xgboost(data = dtrain, nrounds = 100)

# Make predictions
predictions <- predict(xgb_model, newdata = dtest)

```
```{r}
# Convert predicted probabilities to class predictions (assuming binary classification)
predicted_classes <- factor(ifelse(predictions > 0.5, "LumA", "Basal"), levels = c("LumA", "Her2", "LumB", "Basal", "Normal"))

# Calculate accuracytrain
accuracy <- sum(predicted_classes == test_labels_subtype) / length(test_labels_subtype)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

```
