---
title: "RNAseq"
author: "Shira, Moria, Naama"
date: "2023-06-22"
output: html_document
---
## Libraries

```{r setup, include=FALSE}
library(googledrive)
library(ggplot2)
library(readr)
library(googledrive)
library(RColorBrewer)
library(dplyr)
library(httr)
```


## Instructions for running the RNA seq

If it's the first time running the code on your device

```{r}
# drive_auth()
# # clinical data
# url="https://drive.google.com/file/d/1sIeLwzRQvszjkUNsefTtVdhS74Y09zbz/view?usp=sharing"
# drive_download(url,overwrite=TRUE)
# # Rna seq
# url="https://drive.google.com/file/d/1Ecqg0itgMpaVxuy3lbQt_OHQm77JQpif/view?usp=sharing"
# drive_download(url,overwrite=TRUE)
# # mutations
# url="https://drive.google.com/file/d/1165vt_jwnRiXLBHmGyC1XAobGGSPQXRt/view?usp=sharing"
# drive_download(url,overwrite=TRUE)
# 
# url="https://drive.google.com/file/d/1_98OI7FKtb_Ky3lHyh0QwYPt43xYZkF2/view?usp=sharing"
# drive_download(url,overwrite=TRUE)
# rna_seq_data_x<-read.csv(drive_download(url,overwrite=TRUE))
```

brca_tcga_clinical_data.tsv
```{r}
#or read.csv("Breast_cancer_staging_V2.csv")
bc_data <- read.delim("brca_tcga_clinical_data.tsv")
head(bc_data)
```
scaled_x_rna_seq2.csv
```{r}
scaled_data <- read.csv("scaled_x_rna_seq2.csv")
dim(scaled_data)
head(scaled_data)
```
y_mutations
```{r}
y_mutations <- read.csv("y_mutations2.csv")
dim(y_mutations)
head(y_mutations)
```
bc_data_df
```{r}
# Selecting specific columns using subset()
bc_data_df <- subset(bc_data, select = c(Patient.ID,ER.Status.By.IHC,IHC.HER2,PR.status.by.ihc,Overall.Survival.Status))
head(bc_data_df)
```

```{r}
Breast_cancer <- read.delim("brca_tcga_clinical_data.tsv")
```
## pie charts

```{r}
er <- table(Breast_cancer$ER.Status.By.IHC)
myPalette <- brewer.pal(length(er), "Set3") 
pie(er , labels = names(er), border="white", col=myPalette ,main="ER Status")
```

```{r}
her2 <- table(Breast_cancer$IHC.HER2)
myPalette <- brewer.pal(length(her2), "Set3") 
pie(her2 , labels = names(her2), border="white", col=myPalette ,main="HER2 Status")
```

```{r}
pr <- table(Breast_cancer$PR.status.by.ihc)
myPalette <- brewer.pal(length(pr), "Set3") 
pie(pr , labels = names(pr), border="white", col=myPalette ,main="PR Status")
```

```{r, warning=FALSE}
race_counts <- table(Breast_cancer$Race.Category)
myPalette <- brewer.pal(length(race_counts), "Set2") 
pie(race_counts , labels = c("NATIVE","ASIAN","AFRICAN AMERICAN", "CAUCASIAN"), border="white", col=myPalette )

sex_counts <- table(Breast_cancer$Sex)
myPalette <- brewer.pal(length(sex_counts), "Set3") 
pie(sex_counts , labels = names(sex_counts), border="white", col=myPalette )

survival_counts <- table(Breast_cancer$Overall.Survival.Status)
myPalette <- brewer.pal(length(survival_counts), "Set2") 
pie(survival_counts , labels = c("ALIVE","DEAD"), border="white", col=myPalette )

Menopause <- table(Breast_cancer$Menopause.Status)
myPalette <- brewer.pal(length(Menopause), "Set3") 
pie(Menopause , labels = c("Indeterminate","Peri: 6-12 mos","Post : >12 mos","Pre"), border="white", col=myPalette )

```

```{r, warning=FALSE}
Breast_cancer %>% 
    group_by(Diagnosis.Age) %>%
    count() %>%
    #filter(n>10) %>%
    ggplot() +
    geom_col(aes(Diagnosis.Age,n),fill = 'pink')+
    ggtitle("Age Analysis")+
    xlab("Age")+
    ylab("Age Count")

Age<-(Breast_cancer$Diagnosis.Age)

Mutation_Count<-(Breast_cancer$Mutation.Count)

plot(Age,Mutation_Count, ylab = "Mutation Count", xlab = "Age", main = "Mutation_Count vs. Age")
```

```{r}
numeric_bc <- Breast_cancer[, c("Mutation.Count", "Fraction.Genome.Altered", "Diagnosis.Age", "Days.to.Sample.Collection.", "Lymph.Node.s..Examined.Number", "Sample.Initial.Weight")]
# Calculate the correlation matrix
cor_numeric_bc <- cor(numeric_bc, use = "pairwise.complete.obs")
# Subset the correlation matrix for variables with missing values
cor_missing_values <- cor_numeric_bc[, colSums(is.na(numeric_bc)) > 0]

library(corrplot)
corrplot(cor_missing_values, method = "square", type = "upper")
```

Breast_cancer$Neoplasm.American....
Breast_cancer$Patient.Primary.Tumor.Site
Breast_cancer$Patient.Metastatic.Sites
Breast_cancer$Cancer.Type.Detailed
Breast_cancer$Prior.Cancer.Diagnosis.Occurence
Breast_cancer$Surgical.procedure.first
Breast_cancer$Disease.Surgical.Margin.Status
Breast_cancer$Menopause.Status

Breast_cancer$Sex
Breast_cancer$Race.Category

## survival plots
```{r}
#install.packages("survival")
library(survival)
library(survminer)
library(ggpubr)

# Filter the data for GBM samples only
data_bc_filtered <- bc_data
#TP53     
data_bc_filtered$Vital_status <- ifelse(data_bc_filtered$Overall.Survival.Status == "0:LIVING", 0, 1)

# Specify the correct column names for time and status in the Surv() function
fit <- survfit(Surv(Overall.Survival..Months., Vital_status) ~ Sex, data = data_bc_filtered)

# Create a new plot with survival curves for high and low expression
g <- ggsurvplot(fit, data = data_bc_filtered, risk.table = FALSE, pval = TRUE,
                risk.table.y.text = FALSE, risk.table.height = 0.3,
                palette = c("#2E9FDF", "purple"), conf.int = TRUE,
                xlab = "Time from Diagnosis (months)",
                ylab = "Proportion Survival",
                font.x = c(17, "bold", "black"),
                font.y = c(17, "bold", "black"),
                font.tickslab = c(17, "bold", "black"),
                legend.title = "TRERNA1 Expression: ",
                legend = c(0.89, 0.89),
                title = "Survival\n")

# Display the survival plot
g

```

```{r}
tmp <- y_mutations[, c("X", "TP53")]
tmp2<-bc_data[,c("Patient.ID","Overall.Survival.Status","Overall.Survival..Months.")]
tmp2$Overall.Survival.Status<- ifelse(tmp2$Overall.Survival.Status == "0:LIVING", 0, 1)
tail(tmp2)
colnames(tmp) <- c("Patient.ID", "TP53")
head(tmp)
```

```{r}
tp53_data<-(merge(tmp2, tmp, by= "Patient.ID"))

# Specify the correct column names for time and status in the Surv() function
fit <- survfit(Surv(Overall.Survival..Months., Overall.Survival.Status) ~ TP53, data = tp53_data)

# Create a new plot with survival curves for high and low expression
g <- ggsurvplot(fit, data = tp53_data, risk.table = FALSE, pval = TRUE,
                risk.table.y.text = FALSE, risk.table.height = 0.3,
                palette = c("#2E9FDF", "purple"), conf.int = TRUE,
                xlab = "Time from Diagnosis (months)",
                ylab = "Proportion Survival",
                font.x = c(17, "bold", "black"),
                font.y = c(17, "bold", "black"),
                font.tickslab = c(17, "bold", "black"),
                legend.title = "TRERNA1 Expression: ",
                legend = c(0.89, 0.89),
                title = "Survival\n")
                #xlim = c(0, 20), ggtheme = theme_bw()

# Display the survival plot
g

```

```{r}
print(head(tp53_data))
#TP53_data <- merge(tmp2, tmp, by.x = "Patient.ID", by.y = "X")
```
```{r}
# Assuming your data is stored in a data frame called 'data'
# Extract the features (X) and class labels (y)
#scaled_data

RNAseq_data_survival <- merge(scaled_data, subset(bc_data, select = c(Patient.ID,Overall.Survival.Status)), by.x = "X", by.y = "Patient.ID")
RNAseq_data_survival <- na.omit(RNAseq_data_survival)

head(RNAseq_data_survival)
data_rna_survival <- RNAseq_data_survival[,2:56500]
head(colnames(data_rna_survival))
```

## PCA on y_mutation data and scaled rna seq

mereged mutation data
```{r}
mutation_data <- merge(y_mutations, bc_data_df, by.x = "X", by.y = "Patient.ID")
head(mutation_data)
data_x_y<-merge(mutation_data, scaled_data, by.x = "X", by.y = "X")
```

```{r}
head(data_x_y)
data_x_y_pca <- data_x_y[, !(names(data_x_y) %in% c("X", "Overall.Survival.Status", "ER.Status.By.IHC", "IHC.HER2", "PR.status.by.ihc"))]
```

```{r}
standardized_data_XY <- scale(data_x_y_pca[, -ncol(data_x_y_pca)])
pca_result_XY <- prcomp(standardized_data_XY)
pc_scores_XY <- as.data.frame(pca_result_XY$x)
```

```{r}
pc_scores_XY$Overall.Survival.Status <- factor(mutation_data$Overall.Survival.Status)
pc_scores_XY$ER.Status.By.IHC <- factor(mutation_data$ER.Status.By.IHC)
pc_scores_XY$IHC.HER2 <- factor(mutation_data$IHC.HER2)
pc_scores_XY$PR.status.by.ihc <- factor(mutation_data$PR.status.by.ihc)
```

```{r}
ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = Overall.Survival.Status)) +
  geom_point() +
  labs(title = "PCA Plot - Overall Survival Status") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = ER.Status.By.IHC)) +
  geom_point() +
  labs(title = "PCA Plot - ER Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = IHC.HER2)) +
  geom_point() +
  labs(title = "PCA Plot - IHC HER2") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_XY, aes(x = PC1, y = PC2, color = PR.status.by.ihc)) +
  geom_point() +
  labs(title = "PCA Plot - PR Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```
## PCA - on mutation
```{r}
bc_data_df <- subset(bc_data, select = c(Patient.ID,Overall.Survival.Status,ER.Status.By.IHC,IHC.HER2,PR.status.by.ihc))
head(bc_data_df)
mutation_data <- merge(y_mutations, bc_data_df, by.x = "X", by.y = "Patient.ID")
head(mutation_data)
```

```{r}
pca_data_y <- mutation_data[, c(colnames(mutation_data)[2:45])]
standardized_data_y <- scale(pca_data_y[, -ncol(pca_data_y)])
pca_result_y <- prcomp(standardized_data_y)
pc_scores_y <- as.data.frame(pca_result_y$x)
```

```{r}
pc_scores_y$Overall.Survival.Status <- factor(mutation_data$Overall.Survival.Status)
pc_scores_y$ER.Status.By.IHC <- factor(mutation_data$ER.Status.By.IHC)
pc_scores_y$IHC.HER2 <- factor(mutation_data$IHC.HER2)
pc_scores_y$PR.status.by.ihc <- factor(mutation_data$PR.status.by.ihc)
```

```{r}
ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = Overall.Survival.Status)) +
  geom_point() +
  labs(title = "PCA Plot - Overall Survival Status") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = ER.Status.By.IHC)) +
  geom_point() +
  labs(title = "PCA Plot - ER Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = IHC.HER2)) +
  geom_point() +
  labs(title = "PCA Plot - IHC HER2") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores_y, aes(x = PC1, y = PC2, color = PR.status.by.ihc)) +
  geom_point() +
  labs(title = "PCA Plot - PR Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```

```{r}
gene_summaries <- colSums(y_mutations[, -1])
gene_summaries

id_summaries <- rowSums(y_mutations[, -1])
# Create the bar plot
barplot(table(id_summaries), main = "Frequency", xlab = "Number of mutation", ylab = "Count")
```
##ER statuse and rna seq 
```{r}
sub_rna <- subset(bc_data, select = c(Patient.ID,ER.Status.By.IHC))
head(sub_rna)

RNAseq_data_er <- merge(scaled_data, sub_rna, by.x = "X", by.y = "Patient.ID")
#RNAseq_data_er <- as.data.frame(RNAseq_data_er)

RNAseq_data_er <- na.omit(RNAseq_data_er)


head(RNAseq_data_er)
rna_data_rf <- RNAseq_data_er[,2:56499]
er_status <- RNAseq_data_er$ER.Status.By.IHC
er_status <- factor(er_status)



```

```{r}
rna_matrix <- as.matrix(rna_data_rf)

# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 3  # Number of clusters
kmeans_result <- kmeans(rna_matrix, centers = k)

# Get the cluster assignments for each data point
cluster_labels <- kmeans_result$cluster

# Assuming 'Class' column is a factor
class_labels <- RNAseq_data_er$ER.Status.By.IHC

# Visualize the clusters and class labels (optional)
# Example: PCA plot
library(ggplot2)
#library(prcomp)

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results, cluster labels, and class labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels), Class = factor(class_labels))

# Define colors for the class labels
class_colors <- c("Positive" = "blue", "Negative" = "red", "NA" = "green", "Indeterminate" = "orange")

# Plot PCA with clusters and class labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = Class, shape = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using k-means (k = 3)") +
  scale_color_manual(values = class_colors) +
  guides(shape = guide_legend(title = "Cluster"), color = guide_legend(title = "Class"))

```


## knn for progesteron and estrogen
```{r}
# Load required packages
library(caret)
library(class)
library(ggplot2)

# Prepare the data
rna_data <- rna_data_rf # Remove the first column if it contains row names
rna_matrix <- as.matrix(rna_data)  # Convert data to a numeric matrix
rna_data$Class <- as.factor(RNAseq_data_er$ER.Status.By.IHC)  # Convert class column to factor

# Split the data into training and testing sets
set.seed(123)  # Set seed for reproducibility
train_percentage <- 0.7  # Specify the percentage of data for training
train_indices <- createDataPartition(rna_data$Class, p = train_percentage, list = FALSE)
rna_train <- rna_matrix[train_indices, ]
rna_test <- rna_matrix[-train_indices, ]
class_train <- rna_data$Class[train_indices]
class_train <- as.factor(class_train)

# Perform k-NN classification
k <- 4  # Set the number of nearest neighbors
knn_model <- knn(train = rna_train, test = rna_test, cl = class_train, k = k)
knn_predictions <- as.factor(knn_model)

# Evaluate the model
true_labels <- rna_data$Class[-train_indices]  # True class labels for testing set
confusionMatrix(knn_predictions, true_labels)

# Combine the test data with predicted labels
test_data <- data.frame(PC1 = rna_test[, 1], PC2 = rna_test[, 2], Class = true_labels, Prediction = knn_predictions)

# Plot the data
ggplot(test_data, aes(x = PC1, y = PC2, color = Prediction)) +
  geom_point() +
  labs(title = "k-NN Classification on RNA-seq Data",
       x = "PC1", y = "PC2") +
  scale_color_manual(values = c("Positive" = "blue", "Negative" = "red", "Indeterminate" = "green", "NA" = "gray"),
                     labels = c("Positive", "Negative", "Indeterminate", "NA")) +
  theme_minimal()

```
```{r}
sub_rna_pr <- subset(bc_data, select = c(Patient.ID,PR.status.by.ihc))
head(sub_rna_pr)

RNAseq_data_pr <- merge(scaled_data, sub_rna_pr, by.x = "X", by.y = "Patient.ID")
#RNAseq_data_er <- as.data.frame(RNAseq_data_er)

RNAseq_data_pr <- na.omit(RNAseq_data_pr)


head(RNAseq_data_pr)
rna_data_rf_pr <- RNAseq_data_pr[,2:56499]
er_status <- RNAseq_data_pr$ER.Status.By.IHC
er_status <- factor(er_status)

# Load required packages
library(caret)
library(class)
library(ggplot2)

# Prepare the data
rna_data_pr <- rna_data_rf_pr # Remove the first column if it contains row names
rna_matrix_pr <- as.matrix(rna_data_pr)  # Convert data to a numeric matrix
rna_data_pr$Class <- as.factor(RNAseq_data_pr$PR.status.by.ihc)  # Convert class column to factor

# Split the data into training and testing sets
set.seed(123)  # Set seed for reproducibility
train_percentage <- 0.7  # Specify the percentage of data for training
train_indices <- createDataPartition(rna_data$Class, p = train_percentage, list = FALSE)
rna_train <- rna_matrix[train_indices, ]
rna_test <- rna_matrix[-train_indices, ]
class_train <- rna_data$Class[train_indices]
class_train <- as.factor(class_train)

# Perform k-NN classification
k <- 4  # Set the number of nearest neighbors
knn_model <- knn(train = rna_train, test = rna_test, cl = class_train, k = k)
knn_predictions <- as.factor(knn_model)

# Evaluate the model
true_labels <- rna_data$Class[-train_indices]  # True class labels for testing set
confusionMatrix(knn_predictions, true_labels)

# Combine the test data with predicted labels
test_data <- data.frame(PC1 = rna_test[, 1], PC2 = rna_test[, 2], Class = true_labels, Prediction = knn_predictions)

# Plot the data
ggplot(test_data, aes(x = PC1, y = PC2, color = Prediction)) +
  geom_point() +
  labs(title = "k-NN Classification on RNA-seq Data",
       x = "PC1", y = "PC2") +
  scale_color_manual(values = c("Positive" = "blue", "Negative" = "red", "Indeterminate" = "green", "NA" = "gray"),
                     labels = c("Positive", "Negative", "Indeterminate", "NA")) +
  theme_minimal()

```


```{r}
#rna_data_rf<-scale(rna_data_rf)

# Assuming your RNA-seq data is stored in a variable called 'rna_data'

# Remove the first column (X) as it contains row names
#rna_data <- rna_data[, -1]

# Convert data to a numeric matrix
rna_matrix <- as.matrix(rna_data_rf)

# Perform k-means clustering with 3 clusters
set.seed(123)  # For reproducibility
k <- 3  # Number of clusters
kmeans_result <- kmeans(rna_matrix, centers = k)

# Get the cluster assignments for each data point
cluster_labels <- kmeans_result$cluster

# Visualize the clusters (optional)
# Example: PCA plot
library(ggplot2)
#library(prcomp)

# Perform PCA
pca <- prcomp(rna_matrix)

# Create a data frame with PCA results and cluster labels
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(cluster_labels))

# Plot PCA with clusters
ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point() +
  labs(title = "RNA-seq Data Clustering using k-means (k = 3)")

```

```{r}
library(factoextra)
# Generate your data or load your dataset
data <- rna_data_rf[, 1:4]  # Example with the iris dataset

# Estimate the optimal number of clusters
res <- fviz_nbclust(data, kmeans, method = "wss")

# Plot the results
print(res)

```


```{r}
library(caTools)  # Required for the split

set.seed(123)  # Set a seed for reproducibility
split <- sample.split(er_status, SplitRatio = 0.7)  # Split the data into 70% training and 30% testing
rna_train <- rna_data_rf[split, ]
er_train <- er_status[split]
rna_test <- rna_data_rf[!split, ]
er_test <- er_status[!split]

```


## pca on rna seq
```{r}
bc_data_df <- subset(bc_data, select = c(Patient.ID,Overall.Survival.Status,ER.Status.By.IHC,IHC.HER2,PR.status.by.ihc))
head(bc_data_df)
RNAseq_merged <- merge(data, bc_data_df, by.x = "X", by.y = "Patient.ID")
head(RNAseq_merged)
```

```{r}
pca_data <- RNAseq_merged[, c(colnames(RNAseq_merged)[2:56499])]

standardized_data <- scale(pca_data[, -ncol(pca_data)])
pca_result <- prcomp(standardized_data)
pc_scores <- as.data.frame(pca_result$x)

```

```{r}
pc_scores$Overall.Survival.Status <- factor(RNAseq_merged$Overall.Survival.Status)
pc_scores$ER.Status.By.IHC <- factor(RNAseq_merged$ER.Status.By.IHC)
pc_scores$IHC.HER2 <- factor(RNAseq_merged$IHC.HER2)
pc_scores$PR.status.by.ihc <- factor(RNAseq_merged$PR.status.by.ihc)
```

```{r}

ggplot(pc_scores, aes(x = PC1, y = PC2, color = Overall.Survival.Status)) +
  geom_point() +
  labs(title = "PCA Plot - Overall Survival Status") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores, aes(x = PC1, y = PC2, color = ER.Status.By.IHC)) +
  geom_point() +
  labs(title = "PCA Plot - ER Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores, aes(x = PC1, y = PC2, color = IHC.HER2)) +
  geom_point() +
  labs(title = "PCA Plot - IHC HER2") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

ggplot(pc_scores, aes(x = PC1, y = PC2, color = PR.status.by.ihc)) +
  geom_point() +
  labs(title = "PCA Plot - PR Status By IHC") +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```

